---
title: "Reproducible Research Course Project-1"
author: "Aditya Vikram"
date: "2/28/2018"
output: html_document
---

#Loading and Preprocessing the data
Load the dataset "activity.csv" and change the format of Date column to date
```{r}
activity <- read.csv("activity.csv",header = TRUE)
activity[,2] <- as.Date(activity$date)
```

Let's see the structure of our data
```{r}
str(activity)
```

#Mean total number of steps taken per day
Let's first make a histogram of the frequency of steps and see the distribution
```{r}
steps_1 <- with(activity,tapply(steps,date,sum,na.rm=TRUE))
hist(steps_1,col = "orange", xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")
```

Let's see the mean and median of the number of steps taken daily
```{r}
mean(steps_1)
median(steps_1)
```

So the mean of the number of steps taken daily is 9354.23, and median of the number of steps taken daily is 10395.

Let's see the summary of the total steps taken daily
```{r}
summary(steps_1)
```

#Average daily activity pattern
Let's make a time series plot of 5-minute interval(x-axis) and the average number of steps taken, averaged across all days(y-axis).
```{r}
avg_steps<-with(activity,tapply(steps,interval,mean,na.rm=TRUE))
intervals<-unique(activity$interval)
new<-data.frame(cbind(avg_steps,intervals))
plot(new$intervals,new$avg_steps,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Average Steps per Interval")
```

The 5-minute interval with maximum number of steps
```{r}
index<-which.max(new$avg_steps)
new[index,2]
```

The 5-minute interval that contains maximum number of steps is 835.

#Imputing missing values
Strategy: Calculate the average of average steps per day across all dates in the data set (ignoring NA values). Then use the resulting value in place of NAs.

So, let???s create a new dataset that is equal to the original dataset but with the missing data filled in.

Number of missing values in the dataset.
```{r}
sum(is.na(activity$steps))
```

```{r}
index<-which(is.na(activity$steps))
l<-length(index)
steps_avg<-with(activity,tapply(steps,date,mean,na.rm=TRUE))
na<-mean(steps_avg,na.rm = TRUE)
for (i in 1:l) {
        activity[index[i],1]<-na
}
```

Now let's see if all the missing values are filled.
```{r}
sum(is.na(activity$steps))
```

It looks like we have zero NAs which means we have successfully filled all missing data and new dataset looks like as follows:
```{r}
str(activity)
```

Let???s see the histogram of total steps taken each day with the new dataset
```{r}
steps_2<-with(activity,tapply(steps,date,sum,na.rm=TRUE))
hist(steps_2,col = "green",xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")
```

Let???s calculate and see the mean and median of the total steps taken each day
```{r}
mean(steps_2)
median(steps_2)
```
The mean of the total number of steps each day is 10766.19 and the median of the total number of steps taken each day is 10766.19. Notice that both mean and median come out to be equal after filling all the missing values.

#Are there differences in activity patterns between weekdays and weekends?
In this section, we will use dplyr package and we need to load it from the library.
```{r}
library(dplyr)
```

Also, we will need to create a new variable in the dataset named ???day??? that shows the day of the week in terms of weekday or weekend.
```{r}
activity_mod<- mutate(activity, day = ifelse(weekdays(activity$date) == "Saturday" | weekdays(activity$date) == "Sunday", "weekend", "weekday"))
activity_mod$day<-as.factor(activity_mod$day)
str(activity_mod)
```

Now, let???s plot the weekday and weekend data in seperate graphs
```{r}
act_wknd<-subset(activity_mod,as.character(activity_mod$day)=="weekend")
act_wkdy<-subset(activity_mod,as.character(activity_mod$day)=="weekday")
steps_wknd<-with(act_wknd,tapply(steps,interval,mean,na.rm=TRUE))
steps_wkdy<-with(act_wkdy,tapply(steps,interval,mean,na.rm=TRUE))
int_wknd<-unique(act_wknd$interval)
int_wkdy<-unique(act_wkdy$interval)
new_wknd<-data.frame(cbind(steps_wknd,int_wknd))
new_wkdy<-data.frame(cbind(steps_wkdy,int_wkdy))
par(mfrow=c(2,1),mar=c(4,4,2,1))
plot(new_wknd$int_wknd,new_wknd$steps_wknd,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Weekend")
plot(new_wkdy$int_wkdy,new_wkdy$steps_wkdy,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Weekday")
```

##It is clear that average steps over the weekends show higher pattern than that of the weekdays.
